---
title: "Simulation and Monte Carlo Methods for an Absorbing Random Walk"
---

# Simulation and Monte Carlo Methods for an Absorbing Random Walk

Consider a simple random walk with absorbing boundaries at $0$ and $N$:

$$
X_{n+1} =
\begin{cases}
X_n+1, & \text{with probability } p,\\
X_n-1, & \text{with probability } q=1-p,
\end{cases}
$$

where

$$
X_0=i, \qquad 0<N.
$$

The process stops when it reaches either boundary. Define the absorption time

$$
T=\min\{n\geq 0:X_n=0\text{ or }X_n=N\}.
$$

We use simulation to estimate quantities such as

$$
\Pr(X_T=N\mid X_0=i)
$$

and

$$
E[T\mid X_0=i].
$$

The simulation results can then be compared with the exact results obtained analytically.

## 1. Set up the random walk

We consider

$$
N=10,\qquad X_0=4,\qquad p=0.6,\qquad q=0.4.
$$

```{r}
N <- 10
i <- 4
p <- 0.6
q <- 1 - p
```

For this example, the exact probability of absorption at $N$ is

$$
u_i
=
\frac{1-(q/p)^i}{1-(q/p)^N}.
$$

The exact expected time to absorption is

$$
v_i
=
\frac{
N\left[1-(q/p)^i\right]/\left[1-(q/p)^N\right]-i
}{p-q}.
$$

```{r static-walk-output-1}
u_exact <- (1 - (q / p)^i) / (1 - (q / p)^N)

v_exact <- (
  N * (1 - (q / p)^i) / (1 - (q / p)^N) - i
) / (p - q)

u_exact
v_exact
```

For these parameter values,

$$
u_4\approx 0.817,
\qquad
v_4\approx 20.8.
$$

## 2. Simulating one path

We first simulate a single realization of the random walk.

At each step, we generate a random number from a Bernoulli distribution:

* with probability $p$, the walk moves one step to the right;
* with probability $q$, it moves one step to the left.

The simulation stops when the walk reaches either $0$ or $N$.

```{r}
simulate_walk <- function(i, N, p) {
  
  x <- i
  path <- x
  
  while (x > 0 && x < N) {
    
    step <- sample(c(-1, 1), size = 1, prob = c(1 - p, p))
    x <- x + step
    
    path <- c(path, x)
  }
  
  path
}
```

Generate one sample path.

```{r}
set.seed(123)

path <- simulate_walk(i, N, p)

path
```

The final value of the path identifies the absorbing state:

```{r}
tail(path, 1)
```

The number of transitions before absorption is

```{r}
length(path) - 1
```

### Visualising one sample path

```{r}
plot(
  0:(length(path) - 1),
  path,
  type = "o",
  pch = 16,
  xlab = "Step n",
  ylab = expression(X[n]),
  ylim = c(0, N)
)

abline(h = c(0, N), lty = 2)
```

A single simulated path gives one possible history of the process. It does not, however, provide a reliable estimate of a probability or an expected value.

This leads to the Monte Carlo method.

## 3. Monte Carlo estimation of the absorption probability

We now simulate many independent random-walk paths.

For each path, record whether the walk is eventually absorbed at $N$.

Define

$$
I_k=
\begin{cases}
1, & \text{if the $k$th path is absorbed at }N,\\
0, & \text{if the $k$th path is absorbed at }0.
\end{cases}
$$

Then

$$
\widehat{u}_i
=
\frac{1}{M}\sum_{k=1}^M I_k
$$

is the Monte Carlo estimate of

$$
u_i=\Pr(X_T=N\mid X_0=i).
$$

### Simulating many paths

```{r}
simulate_absorption <- function(i, N, p) {
  
  path <- simulate_walk(i, N, p)
  
  final_state <- tail(path, 1)
  absorption_time <- length(path) - 1
  
  c(
    final_state = final_state,
    absorption_time = absorption_time
  )
}
```

For example, simulate $M=10{,}000$ paths.

```{r}
set.seed(123)

M <- 10000

results <- replicate(
  M,
  simulate_absorption(i, N, p)
)

results <- t(results)
```

The Monte Carlo estimate of the probability of absorption at $N$ is

```{r}
u_mc <- mean(results[, "final_state"] == N)

u_mc
```

Compare this with the exact value:

```{r}
c(
  Exact = u_exact,
  Monte_Carlo = u_mc
)
```

The Monte Carlo estimate should be close to the exact probability. It will not usually be exactly equal because it is based on a finite number of simulations.

## 4. Monte Carlo estimation of the expected absorption time

The same simulation can estimate the expected time to absorption.

For each simulated path, record its absorption time $T_k$.

The Monte Carlo estimator is

$$
\widehat{v}_i
=
\frac{1}{M}\sum_{k=1}^M T_k.
$$

Thus, we can estimate $E[T\mid X_0=i]$ by taking the average absorption time over the simulated paths.

```{r}
v_mc <- mean(results[, "absorption_time"])

v_mc
```

Compare the simulation result with the exact value:

```{r}
c(
  Exact = v_exact,
  Monte_Carlo = v_mc
)
```

The two values should be reasonably close.

## 5. How does the Monte Carlo estimate improve?

Monte Carlo estimates become more stable as the number of simulations increases.

We can investigate this by repeating the simulation for different values of $M$.

```{r}
set.seed(123)

M_values <- c(100, 500, 1000, 5000, 10000)

mc_probability <- numeric(length(M_values))

for (j in seq_along(M_values)) {
  
  M <- M_values[j]
  
  final_states <- replicate(
    M,
    tail(simulate_walk(i, N, p), 1)
  )
  
  mc_probability[j] <- mean(final_states == N)
}

data.frame(
  M = M_values,
  Monte_Carlo = mc_probability
)
```

We can also visualise the convergence.

```{r}
plot(
  M_values,
  mc_probability,
  type = "o",
  pch = 16,
  xlab = "Number of simulations M",
  ylab = "Estimated probability",
  ylim = c(0.75, 0.90)
)

abline(h = u_exact, lty = 2)
```

The horizontal line shows the exact probability $u_4$. As $M$ increases, the Monte Carlo estimate tends to become more stable around the exact value.

## 6. Simulation versus Monte Carlo

It is useful to distinguish two related ideas.

**Simulation** generates possible sample paths of the stochastic process. For example, one simulated path shows one possible sequence

$$
X_0,X_1,X_2,\ldots,X_T.
$$

**Monte Carlo estimation** uses many independent simulated paths to estimate a quantity of interest. For example,

$$
\widehat{u}_i
=
\frac{1}{M}
\sum_{k=1}^M
I_k
$$

estimates the probability of absorption at $N$, while

$$
\widehat{v}_i
=
\frac{1}{M}
\sum_{k=1}^M T_k
$$

estimates the expected absorption time.

Thus, simulation provides the data, while the Monte Carlo method uses the simulated data to estimate a probability, expectation, or other quantity.

::: {.callout-note}

## Key Idea

For an absorbing random walk, simulation can reproduce individual sample paths, while Monte Carlo methods use a large number of simulated paths to approximate quantities such as

$$
\Pr(X_T=N\mid X_0=i)
\quad\text{and}\quad
E[T\mid X_0=i].
$$

The exact analytical results provide a benchmark against which the simulation estimates can be checked.

:::

## 7. A complete Monte Carlo function

The simulation can be combined into one function that estimates both quantities.

```{r}
monte_carlo_absorbing_walk <- function(i, N, p, M = 10000) {
  
  results <- replicate(
    M,
    simulate_absorption(i, N, p)
  )
  
  results <- t(results)
  
  probability <- mean(results[, "final_state"] == N)
  expected_time <- mean(results[, "absorption_time"])
  
  c(
    absorption_probability = probability,
    expected_absorption_time = expected_time
  )
}
```

For example,

```{r}
set.seed(123)

monte_carlo_absorbing_walk(
  i = 4,
  N = 10,
  p = 0.6,
  M = 10000
)
```

The output can be compared directly with

```{r}
c(
  absorption_probability = u_exact,
  expected_absorption_time = v_exact
)
```

This gives a computational verification of the analytical results for the absorbing random walk.
:::

