---
title: "Activity: Explore DTMC Behaviour"
format:
  pdf:
    toc: false
number-sections: false
---

### Activity: Explore DTMC Behaviour

Run the Shiny app and experiment with different transition probabilities and initial distributions. Record your observations and submit your answers through the Google Form.

1. **Initial distribution:** Change $\boldsymbol{\pi}^{(0)}$. How does it affect the probabilities over time?

2. **Long-run behaviour:** Do the state probabilities settle down? What do they approach?

3. **Transition matrix:** Change $P$. How does this affect the long-run probabilities?

4. **Convergence:** Can you find a transition matrix for which the probabilities do not settle down? Why?

**Instruction:** Use the app to explore each question. Give a short answer based on your observations. Where possible, include the transition matrix or initial distribution you used.





### Solutions

## Main Findings

1. **Initial distribution:** Different initial distributions affect the short-run probabilities, but for the NCD chain they eventually lead to the same long-run distribution.

2. **Long-run behaviour:** The state probabilities settle down to the stationary distribution

$$
\boldsymbol{\pi}
=
(0.0476,\ 0.1905,\ 0.7619)
$$

when $p=0.8$.

3. **Transition matrix:** Changing the transition probabilities changes the long-run distribution. For example, when $p$ increases from $0.6$ to $0.8$, the long-run probability of State 3 increases from approximately $0.474$ to $0.762$.

4. **Convergence:** The probabilities do not always settle down. A periodic chain can cause the probabilities to oscillate instead of converging.

---

#### 1. Initial distribution

**Question:** Change $\boldsymbol{\pi}^{(0)}$. How does it affect the probabilities over time?

**Answer:** The initial distribution affects the probabilities in the short run. However, for the NCD chain, the effect disappears as $n$ increases. Since the chain is finite, irreducible, and aperiodic,

$$
\boldsymbol{\pi}^{(n)}\longrightarrow\boldsymbol{\pi}
$$

for every initial distribution $\boldsymbol{\pi}^{(0)}$.

For $p=0.8$,

$$
\boldsymbol{\pi}
=
(0.04761905,\ 0.1904762,\ 0.7619048).
$$

Thus, different initial distributions eventually lead to the same long-run distribution.

---

#### 2. Long-run behaviour

**Question:** Do the state probabilities settle down? What do they approach?

**Answer:** Yes. The state probabilities settle down to

$$
\boldsymbol{\pi}
=
(0.04761905,\ 0.1904762,\ 0.7619048).
$$

Thus, in the long run, the probabilities of being in States 1, 2, and 3 are approximately

$$
4.76\%,\qquad 19.05\%,\qquad 76.19\%.
$$

These are the components of the unique stationary distribution.

---

#### 3. Transition matrix

**Question:** Change $P$. How does this affect the long-run probabilities?

**Answer:** Changing the transition probabilities changes the long-run distribution. For example, changing $p$ in the NCD model changes

$$
\boldsymbol{\pi}
=
\left(
\frac{(p-1)^2}{p^2-p+1},
\frac{p-p^2}{p^2-p+1},
\frac{p^2}{p^2-p+1}
\right).
$$

For example:

* If $p=0.6$,

$$
\boldsymbol{\pi}
\approx
(0.211,\ 0.316,\ 0.474).
$$

* If $p=0.8$,

$$
\boldsymbol{\pi}
\approx
(0.0476,\ 0.1905,\ 0.7619).
$$

Thus, changing the transition probabilities can substantially change the long-run probabilities.

---

#### 4. Convergence

**Question:** Can you find a transition matrix for which the probabilities do not settle down? Why?

**Answer:** Yes. For example, consider

$$
P=
\begin{bmatrix}
0&1&0\\
1&0&0\\
0&0&1
\end{bmatrix}.
$$

If the chain starts in State 1, it alternates between States 1 and 2:

$$
(1,0,0)
\rightarrow
(0,1,0)
\rightarrow
(1,0,0)
\rightarrow\cdots
$$

The probabilities do not settle down. This happens because the chain is **periodic**.

This contrasts with the NCD chain, which is aperiodic, so its state probabilities converge to the unique stationary distribution.





## R code: Results for the four questions

The following code uses the same calculation as the app:

$$
\boldsymbol{\pi}^{(n+1)}=\boldsymbol{\pi}^{(n)}P.
$$

It does not calculate the stationary distribution directly. The long-run values are obtained from the simulated probability evolution, as students would see in the app.


```{r}
library(ggplot2)
library(reshape2)
#library(knitr)
library(kableExtra)

# ------------------------------------------------------------
# Function used by the Shiny app to calculate probability evolution
# ------------------------------------------------------------

prob_evolution <- function(P, init, steps = 20) {
  
  init <- init / sum(init)
  
  dist <- matrix(
    0,
    nrow = steps + 1,
    ncol = 3
  )
  
  dist[1, ] <- init
  
  for (i in 2:(steps + 1)) {
    dist[i, ] <- dist[i - 1, ] %*% P
  }
  
  df <- data.frame(
    Step = 0:steps,
    dist
  )
  
  colnames(df)[2:4] <- c(
    "State 1",
    "State 2",
    "State 3"
  )
  
  df
}


# ------------------------------------------------------------
# Plot function using the same style as the Shiny app
# ------------------------------------------------------------

plot_evolution <- function(df, title) {
  
  df_long <- melt(
    df,
    id.vars = "Step",
    variable.name = "State",
    value.name = "Probability"
  )
  
  ggplot(
    df_long,
    aes(
      x = Step,
      y = Probability,
      color = State
    )
  ) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2) +
    ylim(0, 1) +
    ggtitle(title) +
    theme_minimal()
}
```

## 1. Effect of the initial distribution

Use the NCD transition matrix for $p=0.8$:

```{r}
p <- 0.8

P <- matrix(
  c(
    1 - p, p,     0,
    1 - p, 0,     p,
    0,     1 - p, p
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("1", "2", "3"),
    c("1", "2", "3")
  )
)

P
```

The three initial distributions can then be entered in the app as:

```{r}
init1 <- c(1, 0, 0)
init2 <- c(0, 1, 0)
init3 <- c(0, 0, 1)
```

Calculate the probability evolution for each initial state:

```{r}
dist1 <- prob_evolution(P, init1, steps = 20)
dist2 <- prob_evolution(P, init2, steps = 20)
dist3 <- prob_evolution(P, init3, steps = 20)
```

For example, the last 10 distributions for an initial state of 1 are:

```{r}
kable(
  tail(dist1, 10),
  digits = 3,
  align = "c"
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover")
  )
```

Plot the three cases separately, as students would see by changing the initial distribution in the app:

```{r}
plot_evolution(
  dist1,
  "Initial Distribution: State 1"
)

plot_evolution(
  dist2,
  "Initial Distribution: State 2"
)

plot_evolution(
  dist3,
  "Initial Distribution: State 3"
)
```

For a direct comparison of the three initial distributions, we can also plot the probability of each state across the three runs:

```{r}
df1 <- dist1
df1$Initial <- "State 1"

df2 <- dist2
df2$Initial <- "State 2"

df3 <- dist3
df3$Initial <- "State 3"

comparison <- rbind(df1, df2, df3)

comparison_long <- melt(
  comparison,
  id.vars = c("Step", "Initial"),
  variable.name = "State",
  value.name = "Probability"
)

ggplot(
  comparison_long,
  aes(
    x = Step,
    y = Probability,
    color = State
  )
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.5) +
  ylim(0, 1) +
  facet_wrap(~ Initial) +
  ggtitle("Effect of the Initial Distribution") +
  theme_minimal()
```

The three graphs show that the paths are different initially, but they approach the same values.

---

## 2. Long-run behaviour

Starting from State 1, as in the original NCD example:

```{r}
dist <- prob_evolution(
  P,
  init = c(1, 0, 0),
  steps = 20
)

tail(dist, 10)
```

The corresponding graph is:

```{r}
plot_evolution(
  dist,
  "Long-run Evolution of NCD Probabilities"
)
```

To see the approximate limiting values more clearly, increase the number of steps:

```{r}
dist_50 <- prob_evolution(
  P,
  init = c(1, 0, 0),
  steps = 50
)

tail(dist_50, 10)
```

The final row gives approximately

```{r}
tail(dist_50, 1)
```

which is close to

$$
(0.0476,\ 0.1905,\ 0.7619).
$$

This is the same long-run distribution shown in the lecture slides.

---

## 3. Effect of changing the transition matrix

Students can change the transition probabilities in the app. To reproduce the NCD examples, compare $p=0.6$ and $p=0.8$.

```{r}
P_06 <- matrix(
  c(
    0.4, 0.6, 0,
    0.4, 0,   0.6,
    0,   0.4, 0.6
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("1", "2", "3"),
    c("1", "2", "3")
  )
)

P_08 <- matrix(
  c(
    0.2, 0.8, 0,
    0.2, 0,   0.8,
    0,   0.2, 0.8
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("1", "2", "3"),
    c("1", "2", "3")
  )
)

P_06
P_08
```

Starting from State 1:

```{r}
dist_06 <- prob_evolution(
  P_06,
  c(1, 0, 0),
  steps = 20
)

dist_08 <- prob_evolution(
  P_08,
  c(1, 0, 0),
  steps = 20
)
```

Show the final 10 rows for each:

```{r}
kable(
  tail(dist_06, 10),
  digits = 3,
  align = "c"
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center"
  )

kable(
  tail(dist_08, 10),
  digits = 3,
  align = "c"
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

And plot the two cases:

```{r}
plot_evolution(
  dist_06,
  "NCD Evolution: p = 0.6"
)

plot_evolution(
  dist_08,
  "NCD Evolution: p = 0.8"
)
```

The graphs show that changing the transition probabilities changes the long-run probabilities. In particular, increasing $p$ makes State 3 much more likely in the long run.

---

## 4. An example where probabilities do not settle down

Use the following transition matrix:

```{r}
P_periodic <- matrix(
  c(
    0, 1, 0,
    1, 0, 0,
    0, 0, 1
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("1", "2", "3"),
    c("1", "2", "3")
  )
)

P_periodic
```

Start from State 1:

```{r}
dist_periodic <- prob_evolution(
  P_periodic,
  c(1, 0, 0),
  steps = 20
)

dist_periodic
```

The graph is:

```{r}
plot_evolution(
  dist_periodic,
  "Periodic Chain: Probabilities Do Not Settle Down"
)
```

The output alternates between

$$
(1,0,0)
\quad\text{and}\quad
(0,1,0).
$$

Hence, the probabilities do not converge.

The important observation from the graph is that the probability of State 1 alternates between $1$ and $0$, while the probability of State 2 alternates between $0$ and $1$.



